### 2025-02-03 paper/preprint/paper.tex
beautiful

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems involving opaque
models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
tools can aide us in monitoring models and providing affected
individuals with recourse
(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}).

@@

In this work, we build on these findings and introduce a
\textbf{counterfactual training}: a novel approach geared explicitly
towards aligning model representations with desirable explanations.

@@

\section{Counterfactual Training}\label{counterfactual-training}

@@

>>>>

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems involving opaque
models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
tools can aide us in monitoring models and providing recourse to
affected individuals (\citeproc{ref-wachter2017counterfactual}{Wachter,
Mittelstadt, and Russell 2017}).

@@

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with desirable explanations. Our
contributions are as follows:

\begin{itemize}
\tightlist
\item
  We discuss existing related work on improving models and consolidate
  it through the lens of counterfactual explanations
  (Section~\ref{sec-lit}).
\item
  We present our proposed methodological framework that leverages
  faithful counterfactual explanations during the training phase of
  models to achieve the explainability objective
  (Section~\ref{sec-method}).
\item
  Through extensive experiments we demonstrate the counterfactual
  training improve model explainability while maintaining high
  predictive performance. We run ablation studies and grid searches to
  understand how the underlying model components and hyperparameters
  affect outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude that counterfactual training
provides a practical framework for researchers and practitioners
interested in making opaque models more trustworthy
Section~\ref{sec-conclusion}. We also believe that this work serves as
an opportunity for XAI researchers to reevaluate the premise of
improving XAI tools without improving models.

@@

\section{Counterfactual Training}\label{sec-method}

@@

\bibitem[\citeproctext]{ref-augustin2020adversarial}
Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
{``Adversarial Robustness on in-and Out-Distribution Improves
Explainability.''} In \emph{European Conference on Computer Vision},
228--45. Springer.

### 2025-02-03 paper/sections/intro.qmd
beautiful

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing affected individuals with recourse [@wachter2017counterfactual].

@@

Building on these findings, we introduce a **counterfactual training**: a novel approach geared explicitly towards aligning model representations with desirable explanations.

From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.

>>>>

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing recourse to affected individuals [@wachter2017counterfactual].

@@

Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations. Our contributions are as follows:

- We discuss existing related work on improving models and consolidate it through the lens of counterfactual explanations (@sec-lit).
- We present our proposed methodological framework that leverages faithful counterfactual explanations during the training phase of models to achieve the explainability objective (@sec-method).
- Through extensive experiments we demonstrate the counterfactual training improve model explainability while maintaining high predictive performance. We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes. (@sec-experiments).

Despite limitations of our approach discussed in @sec-discussion, we conclude that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy [@sec-conclusion]. We also believe that this work serves as an opportunity for XAI researchers to reevaluate the premise of improving XAI tools without improving models.

### 2025-02-03 paper/sections/related.qmd
streamlined quarto bs
intro basically done
beautiful

@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*. The demonstrate that using these generated counterfactuals during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.

@@

>>>>

@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*. They demonstrate that using these generated counterfactuals during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.

@@

From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.
